---
title: "First Year Project Beautiful"
output: html_document
date: "14-03-2023"
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Packages
```{r}
library(tidyverse)
```

# Reading Data

```{r}
data <- read.csv("Factors Affecting Extinction.csv")
head(data)
```
We have the species, which is our identifier of each bird. Time (extinction time) is a numerical value, as is Paris (nesting pairs). Both size and status are categorical variables with either [L (Large),S (Small)] or [R (Resident),M (Migrant)]

# Making a full model
We make a full model describing extinction time (outcome variable) as a combination of pairs, size and status (predictor variables)
The model will look like this:
$$
\hat{Y} = \hat{\beta_0} + \hat{\beta_1} \times \text{Pairs} + \hat{\beta_2}  \times \text{Size} + \hat{\beta_3} \times \text{Status} + \hat{\beta_4} \times \text{Pairs} \times \text{Size} + \hat{\beta_5} \times \text{Pairs} \times \text{Status} + \hat{\beta_6} \times \text{Status} \times \text{Size} + \hat{\beta_7} \times \text{Pairs} \times \text{Size} \times \text{Status}
$$

```{r}
lm_full <- lm(formula = Time ~ Pairs + Size + Status + Pairs*Size + Pairs * Status + Size*Status + Pairs*Size*Status, data)
summary(lm_full)

```
Calling the summary on the Multiple linear model we just created, we get different measurements. 
Looking at the R-squared, we got a R-Squared value of 0.3398. We also get an adjusted R-squared with a value of 0.2542. 
Both show a weak correlation of predictor variables and outcome variables. 
The p-value is 0.0014 which means this data would occur randomly, approximately 1.4% of the time.
Looking at the summary we see that R has encoded the Size value "Small" as 1, and Status value "Resident" as one. The opposite values are encoded to 0. 
This gives us four different possible model looks:

The first model is where Size = Small and Status = Resident. This model is the same as proposed earlier.
$$
\hat{Y} = \hat{\beta_0} + 
\hat{\beta_1} \times \text{Pairs} + 
\hat{\beta_2} \times \text{Size} + 
\hat{\beta_3} \times \text{Status} + 
\hat{\beta_4} \times \text{Pairs} \times \text{Size} + 
\hat{\beta_5} \times \text{Pairs} \times \text{Status} + 
\hat{\beta_6} \times \text{Status} \times \text{Size} + 
\hat{\beta_7} \times \text{Pairs} \times \text{Size} \times \text{Status}
$$
The second model is the model where Size = Large and Status = Resident. In this model, all Size variabels are encoded to 0, meaning terms including Size are eliminated.

$$
\hat{Y} = \hat{\beta_0} + 
\hat{\beta_1} \times \text{Pairs} + 
\hat{\beta_3} \times \text{Status} + 
\hat{\beta_5} \times \text{Pairs} \times \text{Status}
$$
The third model is where Size = Large and Status = Migrant. Both Status and Size are 0. 

$$
\hat{Y} = \hat{\beta_0} + 
\hat{\beta_1} \times \text{Pairs}
$$
The fourth model is where Size = Small and Status = Migrant. Migrant is eliminated from terms including it. 
$$
\hat{Y} = \hat{\beta_0} + 
\hat{\beta_1} \times \text{Pairs} + 
\hat{\beta_2} \times \text{Size} + 
\hat{\beta_4} \times \text{Pairs} \times \text{Size}
$$

For the next step in the analysis of the dataset, we look at outliers and residuals

# Residual plot of the full model fit



```{r}


```











