---
title: "First_Year_Project_beautiful2"
author: "Andreas"
date: "2023-03-16"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Packages
```{r}
library(tidyverse)
```

# Reading Data

```{r}
data <- read.csv("Factors Affecting Extinction.csv")
head(data)
```

# Scatterplot of dataset with Categorical variables colorized

```{r}

ggplot(data, aes(Pairs, Time,col=Size)) + geom_point() + ggtitle("Time vs Pairs showing Size")+ ylab("Extinction Time")  + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + ylim(c(0,60)) + scale_color_manual(values = c("goldenrod1","dodgerblue2"))

ggplot(data, aes(Pairs, Time,col=Status)) + geom_point() + ggtitle("Time vs Pairs showing Status")+ ylab("Extinction Time")  + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + ylim(c(0,60)) + scale_color_manual(values= c("black", "red2"))

data.for.plot <- data
data.for.plot$Type <- paste(data.for.plot$Size, data.for.plot$Status, sep="")

ggplot(data.for.plot, aes(Pairs, Time,col = Type)) + geom_point() + ggtitle("Time vs Pairs showing Size & Status")+ ylab("Extinction Time")  + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + ylim(c(0,60)) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black", "red2"))

```

# Making a baseline model
```{r}

lm.base <- lm(formula = Time ~ Pairs + Size + Status + Pairs*Size + Pairs*Status + Size*Status + Pairs*Size*Status, data)
summary(lm.base)

```

# Residual plot for baseline model
```{r}
residual.ylim <- ylim(c(-6,6))
standard_res <- rstandard(lm.base)
final_data <- cbind(data, standard_res)
ggplot(final_data) + geom_point(aes(Time, standard_res)) + geom_abline(slope = 0) + residual.ylim + ggtitle("Residual plot for baseline model") + ylab("Standardized Residuals") + xlab("Predicted Values") + theme(plot.title = element_text(hjust = 0.5))

```

# QQ-plot for raw data

```{r}

ggplot(data, aes(sample = Time)) + stat_qq() + stat_qq_line() + ylab("Sample Quantiles") + xlab("Theoretical Quantiles") + ggtitle("QQ-plot for raw data") + theme(plot.title = element_text(hjust = 0.5))

```

# Regression line plot for baseline model with 4 different slopes
```{r}
ggplot(data.for.plot, aes(Pairs, Time,col = Type)) + geom_point() + ggtitle("Time vs Pairs showing Size, Status & Regression Lines")+ ylab("Extinction Time")  + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + ylim(c(0,60)) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black","red2")) + stat_smooth(method = "lm", fullrange=T, level=0)

```

# Scatterplot for different transformations

```{r}
# log transformation
ggplot(data.for.plot, aes(Pairs, log(Time),col = Type)) + geom_point() + ggtitle("log(Time) vs Pairs showing Size & Status")+ ylab("log(Extinction Time)")  + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black", "red2"))

# sqrt transformation
ggplot(data.for.plot, aes(Pairs, sqrt(Time),col = Type)) + geom_point() + ggtitle("sqrt(Time) vs Pairs showing Size & Status")+ ylab("sqrt(Extinction Time)")  + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black", "red2"))

# inverse transformation

ggplot(data.for.plot, aes(Pairs, 1/(Time),col = Type)) + geom_point() + ggtitle("1/Time vs Pairs showing Size & Status")+ ylab("1/(Extinction Time)")  + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black", "red2"))



```
# Transformed lm    

```{r}

lm.log <- lm(formula = log(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs*Status + Size*Status + Pairs*Size*Status, data)
lm.sqrt<- lm(formula = sqrt(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs*Status + Size*Status + Pairs*Size*Status, data)
lm.inv <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs*Status + Size*Status + Pairs*Size*Status, data)
summary(lm.log)
summary(lm.sqrt)
summary(lm.inv)

```

# Residual plot for transformations

```{r}

# log residual
standard_res.log <- rstandard(lm.log)
final_data.log <- cbind(data, standard_res.log)
ggplot(final_data.log) + geom_point(aes(log(Time), standard_res.log)) + geom_abline(slope = 0) + residual.ylim + ggtitle("Residual plot for log transformed model") + ylab("Standardized Residuals") + xlab("Predicted Values") + theme(plot.title = element_text(hjust = 0.5))

# sqrt residual
standard_res.sqrt <- rstandard(lm.sqrt)
final_data.sqrt <- cbind(data, standard_res.sqrt)
ggplot(final_data.sqrt) + geom_point(aes(sqrt(Time), standard_res.sqrt)) + geom_abline(slope = 0) + residual.ylim + ggtitle("Residual plot for baseline sqrt transformed model") + ylab("Standardized Residuals") + xlab("Predicted Values") + theme(plot.title = element_text(hjust = 0.5))

# inverse residual
standard_res.inv <- rstandard(lm.inv)
final_data.inv <- cbind(data, standard_res.inv)

ggplot(final_data.inv) + geom_point(aes(1/Time, standard_res.inv)) + geom_abline(slope = 0) + residual.ylim + ggtitle("Residual plot for inversely transformed model") + ylab("Standardized Residuals") + xlab("Predicted Values") + theme(plot.title = element_text(hjust = 0.5))

```

# QQ-plot for transformed models
```{r}
# log qqplot
ggplot(data, aes(sample = log(Time))) + stat_qq() + stat_qq_line() + ylab("Sample Quantiles") + xlab("Theoretical Quantiles") + ggtitle("QQ-plot for log transformed data") + theme(plot.title = element_text(hjust = 0.5)) 

# sqrt qqplot
ggplot(data, aes(sample = sqrt(Time))) + stat_qq() + stat_qq_line() + ylab("Sample Quantiles") + xlab("Theoretical Quantiles") + ggtitle("QQ-plot for sqrt transformed data") + theme(plot.title = element_text(hjust = 0.5)) 

# inverse qqplot
ggplot(data, aes(sample = 1/(Time))) + ylab("Sample Quantiles") + xlab("Theoretical Quantiles") + ggtitle("QQ-plot for inversely transformed data") + theme(plot.title = element_text(hjust = 0.5)) + stat_qq() + stat_qq_line()



```

# Regression line plot for transformed model with 4 different slopes
Choosing the inverse transformation
```{r}

ggplot(data.for.plot, aes(Pairs, 1/(Time),col = Type)) + geom_point() + ggtitle("1/Time vs Pairs showing Size, Status & Regression Lines")+ ylab("1/(Extinction Time)") + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black","red2")) + stat_smooth(method = "lm", fullrange=T, level=0)
```

# Plotting outliers using cooks distance

Calculating cooks distance for transformed linear model and plotting

```{r}
ckd.inv <- cooks.distance(lm.inv)
mask <- which(ckd.inv > 0.2)
outliers <- data[c(which(ckd.inv > 0.2)),]
data.outlier.plot <- data
data.outlier.plot$name.plot <- "Others"
for (i in 1:3) {
  data.outlier.plot$name.plot[mask[i]] <- outliers$Species[i]  
}

ggplot(lm.inv, aes(seq_along(.cooksd),.cooksd)) + geom_col(aes(fill = data.outlier.plot$name.plot)) + ylab("Cooks distance") + xlab("Index") + guides(fill = guide_legend(title="Birds")) + ggtitle("Birds with significant cooks distance") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_manual(values = c("firebrick2", "mediumpurple", "black", "goldenrod2"))


```

# Scatterplot and residual plot with highlighted cooks distances

```{r}

ggplot(data.outlier.plot) + geom_point(aes(Pairs, 1/Time, col = name.plot)) + scale_color_manual(values = c("firebrick2", "mediumpurple", "black", "goldenrod2")) + ggtitle("Scatterplot for inversely transformed data \nwith cooks distance highlighted") + theme(plot.title = element_text(hjust=0.5))

standard_res.inv <- rstandard(lm.inv)
final_data.inv <- cbind(data.outlier.plot, standard_res.inv)

ggplot(final_data.inv) + geom_point(aes(1/Time, standard_res.inv, col = name.plot)) + geom_abline(slope = 0) + residual.ylim + ggtitle("Residual plot for inversely transformed model \nwith cooks distance highlighted") + ylab("Standardized Residuals") + xlab("Predicted Values") + theme(plot.title = element_text(hjust = 0.5)) + scale_color_manual(values = c("firebrick2", "mediumpurple", "black", "goldenrod2"))


```

# Checking if removing outliers does much to model

```{r}

data.minOutliers <- data[-c(which(ckd.inv > 0.2)),]
lm.inv.minOutliers <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs*Status + Size*Status + Pairs*Size*Status, data.minOutliers)

summary(lm.inv)
summary(lm.inv.minOutliers)

```

```{r}

data.minOutliers.forPlot <- data.for.plot[-c(which(ckd.inv > 0.2)),]

ggplot(data.minOutliers.forPlot, aes(Pairs, 1/(Time), col=Type)) + geom_point() + ggtitle("1/Time vs Pairs showing Size, Status & Regression Lines minus outliers")+ ylab("1/(Extinction Time)") + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + ylim(c(0,1)) + stat_smooth(method = "lm", fullrange=T, level=0) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black","red2")) 

ggplot(data.for.plot, aes(Pairs, 1/(Time), col=Type)) + geom_point() + ggtitle("1/Time vs Pairs showing Size, Status & Regression Lines")+ ylab("1/(Extinction Time)") + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + ylim(c(0,1))+ stat_smooth(method = "lm", fullrange=T, level=0) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black","red2")) 



```

Both adjusted R-squared and R-squared are bad for all transformations.

# Scatterplots of transformation of independent variable

```{r}

# log independent variable transformation
ggplot(data.for.plot, aes(Pairs, log(Time),col = Type)) + geom_point() + ggtitle("Time vs log(Pairs) showing Size & Status")+ ylab("Extinction Time")  + xlab("log(Nr. of nesting pairs)") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black", "red2"))

# sqrt independent variable transformation
ggplot(data.for.plot, aes(Pairs, sqrt(Time),col = Type)) + geom_point() + ggtitle("Time vs sqrt(Pairs) showing Size & Status")+ ylab("Extinction Time")  + xlab("sqrt(Nr. of nesting pairs)") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black", "red2"))

# inverse independent variable transformation

ggplot(data.for.plot, aes(Pairs, 1/(Time),col = Type)) + geom_point() + ggtitle("Time vs 1/Pairs showing Size & Status")+ ylab("Extinction Time")  + xlab("1/(Nr. of nesting pairs)") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black", "red2"))



```

# Transformation of independent variable

```{r}

lm.ind.log <- lm(formula = Time ~ log(Pairs) + Size + Status + log(Pairs)*Size + log(Pairs)*Status + Size*Status + log(Pairs)*Size*Status, data)
lm.ind.sqrt<- lm(formula = Time ~ sqrt(Pairs) + Size + Status + sqrt(Pairs)*Size + sqrt(Pairs)*Status + Size*Status + sqrt(Pairs)*Size*Status, data)
lm.ind.inv <- lm(formula = Time ~ 1/(Pairs) + Size + Status + (1/(Pairs))*Size + (1/(Pairs))*Status + Size*Status + (1/(Pairs))*Size*Status, data)
summary(lm.ind.log)
summary(lm.ind.sqrt)
summary(lm.ind.inv)


```

# Comparison models using anova to simplify model if possible


```{r}

m1 <- lm(formula = 1/(Time) ~ Pairs, data)
m2 <- lm(formula = 1/(Time) ~ Pairs + Size, data)
m3 <- lm(formula = 1/(Time) ~ Pairs + Size + Status, data)
m4 <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Status, data)
m5 <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Status + Pairs*Size, data)
m6 <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Status + Pairs*Size  + Pairs*Size*Status, data)

summary(m1)$adj.r.squared
summary(m2)$adj.r.squared
summary(m3)$adj.r.squared
summary(m4)$adj.r.squared
summary(m5)$adj.r.squared
summary(m6)$adj.r.squared

anova(m1,m2,m3,m4,m5,m6)

```

```{r}
m9 <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Status  + Pairs*Size, data)
summary(m9)
```
 Anova returns a p-value for which if p > 0.5 we prefer the more simple model
```{r}

#ohterway..

m1 <- lm(formula = 1/(Time) ~ Pairs, data)
m2 <- lm(formula = 1/(Time) ~ Pairs + Size, data)
m3 <- lm(formula = 1/(Time) ~ Pairs + Size + Pairs*Size, data)
m4 <- lm(formula = 1/(Time) ~ Pairs + Status, data)
m5 <- lm(formula = 1/(Time) ~ Pairs + Status + Pairs * Status, data)
m6 <- lm(formula = 1/(Time) ~ Pairs + Size + Status, data)
m7 <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Size, data)
m8 <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Status, data)
m9 <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Status  + Pairs*Size, data)
m10<- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs*Status + Pairs*Size*Status, data)

anova(m10, m1)
anova(m10, m2)
anova(m10, m3)
anova(m10, m4)
anova(m10, m5)
anova(m10, m6)
anova(m10, m7)
anova(m10, m8)
anova(m10, m9)

```



# Final model

#####shouldn't we work with the data withour outliers???

```{r}
data.final <- cbind(data.for.plot, predict(m9))

vSR.slope <- (-0.09340) + (1*(-0.04936)) + (1*0.06355)
vSR.intercept <- 0.67505 + (1*0.37217) + (1*-0.35078)
vLR.slope <- (-0.09340) + (0*(-0.04936)) + (1*0.06355)
vLR.intercept <- 0.67505 + (0*0.37217) + (1*-0.35078)
vSM.slope <- (-0.09340) + (1*(-0.04936)) + (0*0.06355)
vSM.intercept <- 0.67505 + (1*0.37217) + (0*-0.35078)
vLM.slope <- (-0.09340) + (0*(-0.04936)) + (0*0.06355)
vLM.intercept <- 0.67505 + (0*0.37217) + (0*-0.35078)

ggplot(data.for.plot, aes(Pairs, 1/(Time), col=Type)) + geom_point() + ggtitle("1/Time vs Pairs showing Size, Status & \nRegression Lines for full model")+ ylab("1/(Extinction Time)") + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + ylim(c(0,1))+ stat_smooth(method = "lm", fullrange=T, level=0) + scale_color_manual(values = c("goldenrod1","dodgerblue2", "black","red2")) 


my_cols <- c("goldenrod1","dodgerblue2", "black","red2")
ggplot(data.final, aes(Pairs, 1/(Time), col=Type)) + geom_point() + ggtitle("1/Time vs Pairs showing Size, Status & \nRegression Lines for reduced model")+ ylab("1/(Extinction Time)") + xlab("Nr. of nesting pairs") + theme(plot.title = element_text(hjust = 0.5))+ xlim(c(0,12)) + ylim(c(0,1)) + scale_color_manual(values = my_cols) + 
  geom_abline(slope = vSR.slope, intercept = vSR.intercept, col=my_cols[4], linewidth = 1) +
  geom_abline(slope = vLR.slope, intercept = vLR.intercept, col=my_cols[2], linewidth = 1) +
  geom_abline(slope = vSM.slope, intercept = vSM.intercept, col=my_cols[3], linewidth = 1) +
  geom_abline(slope = vLM.slope, intercept = vLM.intercept, col=my_cols[1], linewidth = 1)



```


