---
title: "First Year Project Beautiful"
output: html_document
date: "14-03-2023"
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Packages
```{r}
library(tidyverse)
```

# Reading Data

```{r}
data <- read.csv("Factors Affecting Extinction.csv")
head(data)
```
2.1

We have the species, which is our identifier of each bird. Time (extinction time) is a numerical value, as is Paris (nesting pairs). Both size and status are categorical variables with either [L (Large),S (Small)] or [R (Resident),M (Migrant)]

# Making a full model
We make a full model describing extinction time (outcome variable) as a combination of pairs, size and status (predictor variables)
The model will look like this:
$$
\hat{Y} = \hat{\beta_0} + \hat{\beta_1} \times \text{Pairs} + \hat{\beta_2}  \times \text{Size} + \hat{\beta_3} \times \text{Status} + \hat{\beta_4} \times \text{Pairs} \times \text{Size} + \hat{\beta_5} \times \text{Pairs} \times \text{Status} + \hat{\beta_6} \times \text{Status} \times \text{Size} + \hat{\beta_7} \times \text{Pairs} \times \text{Size} \times \text{Status}
$$

```{r}
lm_full <- lm(formula = Time ~ Pairs + Size + Status + Pairs*Size + Pairs * Status + Size*Status + Pairs*Size*Status, data)
summary(lm_full)

lm_full2 <- lm(formula = Time ~ Pairs*Size*Status, data)
summary(lm_full2)


```
Calling the summary on the Multiple linear model we just created, we get different measurements. 
Looking at the R-squared, we got a R-Squared value of 0.3398. We also get an adjusted R-squared with a value of 0.2542. 
Both show a weak correlation of predictor variables and outcome variables. 
The p-value is 0.0014 which means this data would occur randomly, approximately 1.4% of the time.
Looking at the summary we see that R has encoded the Size value "Small" as 1, and Status value "Resident" as one. The opposite values are encoded to 0. 
This gives us four different possible model looks:

The first model is where Size = Small and Status = Resident. This model is the same as proposed earlier.
$$
\hat{Y} = \hat{\beta_0} + 
\hat{\beta_1} \times \text{Pairs} + 
\hat{\beta_2} \times \text{Size} + 
\hat{\beta_3} \times \text{Status} + 
\hat{\beta_4} \times \text{Pairs} \times \text{Size} + 
\hat{\beta_5} \times \text{Pairs} \times \text{Status} + 
\hat{\beta_6} \times \text{Status} \times \text{Size} + 
\hat{\beta_7} \times \text{Pairs} \times \text{Size} \times \text{Status}
$$
The second model is the model where Size = Large and Status = Resident. In this model, all Size variabels are encoded to 0, meaning terms including Size are eliminated.

$$
\hat{Y} = \hat{\beta_0} + 
\hat{\beta_1} \times \text{Pairs} + 
\hat{\beta_3} \times \text{Status} + 
\hat{\beta_5} \times \text{Pairs} \times \text{Status}
$$
The third model is where Size = Large and Status = Migrant. Both Status and Size are 0. 

$$
\hat{Y} = \hat{\beta_0} + 
\hat{\beta_1} \times \text{Pairs}
$$
The fourth model is where Size = Small and Status = Migrant. Migrant is eliminated from terms including it. 
$$
\hat{Y} = \hat{\beta_0} + 
\hat{\beta_1} \times \text{Pairs} + 
\hat{\beta_2} \times \text{Size} + 
\hat{\beta_4} \times \text{Pairs} \times \text{Size}
$$

For the next step in the analysis of the dataset, we look at outliers and residuals



*************

## I would say 4 regression lines? it is still the same model, not 4 different models. 







2.1.2

Analyzing assumptions of linear regression


```{r}
ggplot(data, aes(Pairs, Time)) + geom_point()

qqnorm(data$Time, main="Normal Q-Q plot for standarized residuals")
qqline(data$Time)

standard_res <- rstandard(lm_full)
final_data <- cbind(data, standard_res)
ggplot(final_data) + geom_point(aes(Time, standard_res)) + geom_abline(slope = 0)
plot((final_data$Time), standard_res, ylab='Standardized Residuals', xlab='Predicted values', main="Spread-level plot of standardised residuals") 
abline(0, 0)


```


1) Normality. Q-Q plot shows big deviations on each side. Pag 60

2) Homoscedasticity and linearity. Spread-level plot.
    Shows a clear non-linear pattern.



2.2

2.2.1
2.2.2
Transformations of variables 

###### R2, q-q plot and spead-level plot for each transformation of the outcome 'Time'

```{r}
lm_full <- lm(formula = Time ~ Pairs + Size + Status + Pairs*Size + Pairs*Status + Size*Status + Pairs*Size*Status, data)
lm_full.log <- lm(formula = log(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs * Status + Size*Status + Pairs*Size*Status, data)
lm_full.sqrt <- lm(formula = sqrt(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs * Status + Size*Status + Pairs*Size*Status, data)
lm_full.inverse <- lm(formula = 1/Time ~ Pairs + Size + Status + Pairs*Size + Pairs * Status + Size*Status + Pairs*Size*Status, data)

summary(lm_full)$adj.r.squared
summary(lm_full.log)$adj.r.squared
summary(lm_full.sqrt)$adj.r.squared
summary(lm_full.inverse)$adj.r.squared


qqnorm(log(data$Time),main="Q-Q plot for log(time)")
qqline(log(data$Time))
qqnorm(sqrt(data$Time),main="Q-Q plot for sqrt(time)")
qqline(sqrt(data$Time))
qqnorm(1/(data$Time),main="Q-Q plot for 1/time")
qqline(1/(data$Time))

plot(log(final_data$Time), standard_res, ylab='Standardized Residuals', xlab='Predicted values', main="Log(t)") 
abline(0, 0)
plot(sqrt(final_data$Time), standard_res, ylab='Standardized Residuals', xlab='Predicted values', main = "sqrt(t)") 
abline(0, 0)
plot(1/(final_data$Time), standard_res, ylab='Standardized Residuals', xlab='Predicted values', main="1/time")
abline(0, 0)
```

We decided to keep working with the transformed variable log(Time). New model:

```{r}
lm_full_transformed <- lm(formula = log(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs * Status + Size*Status + Pairs*Size*Status, data)
summary(lm_full_transformed)$adj.r.squared

lm_full_inverse <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs * Status + Size*Status + Pairs*Size*Status, data)
summary(lm_full_inverse)$adj.r.squared
```


2.2.3

#Outliers

```{r}
ckd_inv <- cooks.distance(lm_full_inverse)
ckd_log <- cooks.distance(lm_full_transformed)

#sub_ckd <- subset(ckd_inv, ckd_ inv< 0.2)

index <- which(ckd_inv > 0.2)
index.log <- which(ckd_log > 0.2)

index
plot(ckd_inv, ylim = c(0,0.4))
plot(ckd_log, ylim = c(0,0.4))
#plot(sub_ckd, ylim = c(0,0.15))
```

????? in log transf non of them is significantly bigger than the others 


If we remove three outliers in inverse.........

```{r}
data_ckd <- data[-c(index),]
data_ckd_log <- data[-c(index.log),]

lm_full_inverse_ckd <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs*Status + Size*Status + Pairs*Size*Status, data_ckd)
#summary(lm_full_inverse_ckd)#$adj.r.squared
#summary(lm_full_inverse)#$adj.r.squared


lm_full_log_ckd <- lm(formula = log(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs*Status + Size*Status + Pairs*Size*Status, data_ckd_log)
#summary(lm_full_log_ckd)#$adj.r.squared
#summary(lm_full_transformed)


summary(lm_full_inverse_ckd)$adj.r.squared
summary(lm_full_inverse)$adj.r.squared
summary(lm_full_log_ckd)$adj.r.squared
summary(lm_full_transformed)$adj.r.squared


```
```{r}
data_ckd <- data[-c(index),]

lm_final <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs*Status, data_ckd)
summary(lm_final)

chchaaaaaaaaaaa<n
```


2.2.4

#Transforming the independent variable (N of pairs).... dead end 

```{r}
lm_full <- lm(formula = Time ~ Pairs + Size + Status + Pairs*Size + Pairs * Status + Size*Status + Pairs*Size*Status, data)
lm_full.logpair <- lm(formula = Time ~ log(Pairs) + Size + Status + log(Pairs)*Size + log(Pairs) * Status + Size*Status + log(Pairs)*Size*Status, data)
lm_full.sqrtpair <- lm(formula = Time ~ sqrt(Pairs) + Size + Status + sqrt(Pairs)*Size + sqrt(Pairs) * Status + Size*Status + sqrt(Pairs)*Size*Status, data)
lm_full.inversepair <- lm(formula = Time ~ 1/Pairs + Size + Status + 1/Pairs*Size + 1/Pairs * Status + Size*Status + 1/Pairs*Size*Status, data)

summary(lm_full)$r.squared
summary(lm_full.logpair)$r.squared
summary(lm_full.sqrtpair)$r.squared
summary(lm_full.inversepair)$r.squared

```


2.3
Nested models and variable contribution

```{r}

m1 <- lm(formula = 1/(Time) ~ Pairs, data_ckd)
m2 <- lm(formula = 1/(Time) ~ Pairs + Size, data_ckd)
m3 <- lm(formula = 1/(Time) ~ Pairs + Size + Pairs*Size, data_ckd)
m4 <- lm(formula = 1/(Time) ~ Pairs + Status, data_ckd)
m5 <- lm(formula = 1/(Time) ~ Pairs + Status + Pairs * Status, data_ckd)
m6 <- lm(formula = 1/(Time) ~ Pairs + Size + Status, data_ckd)
m7 <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs*Status, data_ckd)     
m8 <- lm(formula = 1/(Time) ~ Pairs + Size + Status + Pairs*Size + Pairs*Status + Pairs*Size*Status + Size*Status, data_ckd)
m9 <- lm(formula = 1/(Time) ~ Pairs + Pairs*Size*Status - Size*Status+Size + Status + Pairs*Size + Pairs*Status , data_ckd)

an <- anova(m1, m2, m3, m4, m5, m6, m7, m8, m9)
#an <- anova(m9, m1)

print(an)
```


```{r}
summary(m1)$adj.r.squared
summary(m2)$adj.r.squared
summary(m3)$adj.r.squared
summary(m4)$adj.r.squared
summary(m5)$adj.r.squared
summary(m6)$adj.r.squared
summary(m7)$adj.r.squared
summary(m8)$adj.r.squared
summary(m9)$adj.r.squared

summary(m7)

```

